Enhancing Adversarial Robustness for Deep Metric Learning

### 2021-10-14 appendix.tex
svg
tex

>>>>

	\item [ICS] should investigate other metric learning loss functions and see
		how they deal with hierarchy. e.g. MS, LSS, GLift, quadrup, etc.

	\item [GA] Also scale the $H_D-H_S$ part?
		(1) $$ H_\mathsf{GA} \leftarrow (H_\mathsf{D} + U - H_\mathsf{S}) * nL $$
		(2) $$ H_\mathsf{GA} \leftarrow H_\mathsf{D} + U * nL $$
		(3) pure GA for ablation $$ H_\mathsf{GA} \leftarrow U * nL $$
		interpret with lessons learned from HM matrix.

### 2021-10-14 robdml.tex
svg
tex

Given an anchor image $X_a$, we sample a positive image $X_p^S$ (in the same
class as the anchor) and a negative image $X_n^S$ (in different class as the
anchor) with a certain sampling strategy (\eg, semi-hard~\cite{facenet}).
%
% For convenience we call this triplet as ``source triplet''.
%
Then its \emph{hardness} is defined as
%
\begin{equation}
%
H(X_a^S,X_p^S,X_n^S)=d_\phi(X_a,X_p^S)-d_\phi(X_a,X_n^S),
%
\end{equation}
%
which is an internal
part of the triplet loss.
%
For convenience we call this as ``source hardness'', denoted as $H_S$.

@@

Instead of directly maximizing the source hardness, we propose to stop the
maximization process when the adversarial examples reach a certain level of
hardness in order to avoid model collapse.
%
Based on the same anchor image $X_a$, we sample another positive image $X_p^D$
and negative image $X_n^D$ with a different triplet sampling strategy, and
we call its hardness as ``destination hardness'', denoted as $H_D$.

Then, we want change (increase) the hardness level from $H_S$ into $H_D$, by
finding adversarial examples for $(X_a, X_p^S, X_n^S)$.
%
Note, it is not expected to lower the hardness during the attack, namely
$E[H_D]$ should be greater or equal to $E[H_S]$.
%
We denote the the hardness of adversarial examples as $\tilde{H}_S=H(X_a{+}r_a,
X_p^S{+}r_p, X_n^S{+}r_n)$.
%
This is named as \emph{Hardness Manipulation} (HM), which is implemented as
follows:
%
\begin{equation}
	%
	\bar{r}_a, \bar{r}_p, \bar{r}_n = \argmin_{r_a,r_p,r_n} \|\min(0,
	\tilde{H}_S-H_D)\|^2.
	%
	\label{eq:hm}
	%
\end{equation}
%
The $\min(0,\cdot)$ part in \cref{eq:hm} truncates the gradient when $H_S>H_D$,
automatically stopping the optimization.
%
This is because $H_S$ will be reduced once greater than $H_D$ in standard Mean
Square Error.
%
The optimization problem can be solved by projected gradient
descent~\cite{madry}.
%
And the resulting adversarial examples are used for adversarially training the
DML model:
%
$L_\text{T}(\phi(X_a+\bar{r}_a), \phi(X_p^S+\bar{r}_p),
\phi(X_n^S+\bar{r}_n))$.

@@

During the training process, the strength of adversary will gradually decrease,
because the expectation of the destination hardness $E[H_D]$ is expected to
decrease with a model being trained.

@@

For cases in upper triangular of \cref{tab:desth}, $E[H_S] \leqslant E[H_D]$,
which means most triplets will be turned into adversarial example by HM
to increase the hardness.
%
Although most of such cases end up with model collapse, the preliminary 
result suggests that HM from random to semihard is effective in significantly
improve the robustness without model collapse, at an expected cost on the R@1
performance on benign examples.

For cases in lower triangular of \cref{tab:desth}, $E[H_S] \geqslant E[H_D]$,
which means most tirplet will not be turned into adversarial according to
\cref{eq:hm}.
%
In such cases, the model will not be able to effectively learn from a strong
adversary, and hence will behave as if no defense is equipped.
%
As an exception, HM from softhard to semihard can also lead to relatively high
adversarial robustness.
%
Due to the high variance of hardness of softhard sampling strategy, the
hardness of a notable portion of triplets will be less than that of semihard,
and hence HM will create some adversarial examples for training.
%
In other words, HM in this case can be interpreted as mixing some unperturbed
softhard triplets and some semihard adversarial triplets.

>>>>

\cref{fig:hm}

Given an anchor image $X_a$, we sample a positive image $X_p^S$ (in the same
class as the anchor) and a negative image $X_n^S$ (in different class as the
anchor) with a certain sampling strategy (\eg, semi-hard~\cite{facenet}).
%
% For convenience we call this triplet as ``source triplet''.
%
Then its \emph{hardness} is defined as
%
\begin{equation}
%
H(X_a^S,X_p^S,X_n^S)=d_\phi(X_a,X_p^S)-d_\phi(X_a,X_n^S),
%
\end{equation}
%
which is an internal
part of the triplet loss.
%
For convenience we call this as ``source hardness'', denoted as $H_\mathsf{S}$.

@@

Instead of directly maximizing the source hardness, we propose to stop the
maximization process when the adversarial examples reach a certain level of
hardness in order to avoid model collapse.
%
Based on the same anchor image $X_a$, we sample another positive image $X_p^D$
and negative image $X_n^D$ with a different triplet sampling strategy, and
we call its hardness as ``destination hardness'', denoted as $H_\mathsf{D}$.

Then, we want change (increase) the hardness level from $H_\mathsf{S}$ into $H_\mathsf{D}$, by
finding adversarial examples for $(X_a, X_p^S, X_n^S)$.
%
Note, it is not expected to lower the hardness during the attack, namely
$E[H_\mathsf{D}]$ should be greater or equal to $E[H_\mathsf{S}]$.
%
We denote the the hardness of adversarial examples as $\tilde{H}_S=H(X_a{+}r_a,
X_p^S{+}r_p, X_n^S{+}r_n)$.
%
This is named as \emph{Hardness Manipulation} (HM), which is implemented as
follows:
%
\begin{equation}
	%
	\bar{r}_a, \bar{r}_p, \bar{r}_n = \argmin_{r_a,r_p,r_n} \|\min(0,
	\tilde{H}_\mathsf{S}-H_\mathsf{D})\|^2.
	%
	\label{eq:hm}
	%
\end{equation}
%
The $\min(0,\cdot)$ part in \cref{eq:hm} truncates the gradient when $H_\mathsf{S}>H_\mathsf{D}$,
automatically stopping the optimization.
%
This is because $H_\mathsf{S}$ will be reduced once greater than $H_\mathsf{D}$ in standard Mean
Square Error.
%
The optimization problem can be solved by projected gradient
descent~\cite{madry}.
%
And the resulting adversarial examples are used for adversarially training the
DML model:
%
$L_\text{T}(\phi(X_a+\bar{r}_a), \phi(X_p^S+\bar{r}_p),
\phi(X_n^S+\bar{r}_n))$.

\oo{uses gradient for maximization. Maxim is $HM(H_\mathsf{S}, 2)$.}

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{gaillust.pdf}
	\caption{Illustration of Gradual Adversary.}
	\label{fig:ga}
\end{figure}

\cref{fig:ga}

During the training process, the strength of adversary will gradually decrease,
because the expectation of the destination hardness $E[H_\mathsf{D}]$ is expected to
decrease with a model being trained.

@@

For cases in upper triangular of \cref{tab:desth}, $E[H_\mathsf{S}] \leqslant E[H_\mathsf{D}]$,
which means most triplets will be turned into adversarial example by HM
to increase the hardness.
%
Although most of such cases end up with model collapse, the preliminary 
result suggests that HM from random to semihard is effective in significantly
improve the robustness without model collapse, at an expected cost on the R@1
performance on benign examples.

For cases in lower triangular of \cref{tab:desth}, $E[H_\mathsf{S}] \geqslant E[H_\mathsf{D}]$,
which means most tirplet will not be turned into adversarial according to
\cref{eq:hm}.
%
In such cases, the model will not be able to effectively learn from a strong
adversary, and hence will behave as if no defense is equipped.
%
As an exception, HM from softhard to semihard can also lead to relatively high
adversarial robustness.
%
Due to the high variance of hardness of softhard sampling strategy, the
hardness of a notable portion of triplets will be less than that of semihard,
and hence HM will create some adversarial examples for training.
%
In other words, HM in this case can be interpreted as mixing some unperturbed
softhard triplets and some semihard adversarial triplets.
